---
title: "A Very Brief R Introduction"
output:
  html_document:
    toc: yes
    df_print: kable
    highlight: tango
#    includes:
#      in_header: wcag-overrides.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# R in Week One

The goal for this document is to introduce you to R documentation and some ways to solve your R problems. It will also include some information on plotting and on exporting your notebooks to HTML and PDF formats.

## Starting with R

### Downloading R
To start with, you need R! For a great walk-through of how to download R and RStudio (my preferred IDE), see this section of [Hands-On Programming with R.](https://rstudio-education.github.io/hopr/starting.html)

If you are in the in-person section, R is available to use on the school computers. You can use the computers in the lab for your project and lab work.

### R Resources
You will need to become accustomed to using [R documentation;](https://www.rdocumentation.org/) no one can ever remember everything that can be done with R! You will often need to look things up.

[Just Enough R's](https://benwhalley.github.io/just-enough-r/) stated goal is "to teach you just enough R to be confident to explore your data." While I hope you will learn more than just enough, this is a great starting point if you're new to using R.

[YaRrr! The Pirate’s Guide to R](https://bookdown.org/ndphillips/YaRrr/) is a fun free book with a lot of great examples.

[R for Data Science](https://r4ds.hadley.nz/) is fantastic as well. Its focus is more on, well, data science than econometrics, but the Explore, Wrangle, Program, and Communicate sections are well worth checking out.

[RStudio produces cheat sheets](https://www.rstudio.com/resources/cheatsheets/) that are easy-to-follow resources on various topics, from data tidying to data visualization.

[StackOverflow](https://stackoverflow.com/questions/tagged/r) is great. If you have a problem, chances are someone has had something similar happen to them and then asked about it on StackOverflow!

Being able to problem solve when coding is an essential skill. It’s a rare day when I don’t get an error message. Become accustomed to looking at and understanding those error messages, and looking at the R documentation to figure out what could be going wrong.

If you are having R problems, I will always ask you one thing that you’ve done to try to solve it on your own.

## Reading Data into R

There are many ways to read data into R. Here I will cover some of them; for details on how to import data from even more file types, [see here.](https://www.datacamp.com/tutorial/r-data-import-tutorial)

### Importing CSV Data
To read in .csv files, you can use the following:

name = read.csv('Path where your .csv file is stored/File Name.csv').

```{r}
#df = read.csv('/Users/jessicaperrigan/Desktop/ECON-8300/FAKE_FILE_NAME.csv')
```

If I tried to run this I'd get an error because there isn't a .csv file named "FAKE_FILE_NAME.csv" in that folder, but all you need to do is replace that with a real file name!

For more information on read.csv(), you can use "?read.csv" (as shown below; to run, remove the #). You can also go to the [R documentation.](https://www.rdocumentation.org/packages/utils/versions/3.6.2/topics/read.table)

```{r}
?read.csv
```

### Importing Data from GitHub

Importing a .csv file from Github is done as the same way as reading a .csv file that you saw above, but this time you're using the URL from Github.

1. Go to the github repository link where you have the CSV file

2. Click on the raw option present on the top right of the data. 

3. This will open a new window in the browser.

4. The link should be like https://raw.githubusercontent.com/...

5. You will use this link in the ```read.csv()``` function. 

```{r}
beer <- read.csv('https://raw.githubusercontent.com/SepAlex/AlexSepe-Project-/main/Beers%20and%20Breweries%2C%20a%20Data%20Analysis/Beers.csv')
```

### Importing .xlsx Data
You can use the read.table() option from R base (it's the same thing the read.csv() runs on), or you can use a special package called "readxl."

To install readxl, you can use ```install.packages()```. To use, delete the "#".

```{r}
#install.packages('readxl')
```

If you also have readxl installed, you can use ```library()``` to make it available.

```{r}
library(readxl)
```

Then the format to read in .xlsx files is ```name = read_excel('Path where your .xlsx file is stored/File Name.xlsx')```.

### Importing Data from Packages

This is how you will import the data you need for homework assignments. To import data from a package, you first need to install the package if it isn't already.

You can install a package by typing the function ```install.package(“package_name”)```. 

To install the Wooldridge package, you can use ```install.packages()```. To use, delete the "#".

```{r}
#install.packages('wooldridge')
```

If you aren't sure if the package is installed, you can use ```installed.packages()``` to get a full list. Or, if you would prefer to check whether a specific package is installed, you can do the following (just replace the "wooldridge" and "AER" package names with the one(s) you're looking for):

```{r}
packages_to_find <- c("wooldridge", "AER") 

ip <- as.data.frame(installed.packages()[,c(1, 3)]) 
rownames(ip) = NULL 

ip[is.element(ip$Package, packages_to_find), ]
```
Once you've installed the package or made sure it's installed, it's simple to import data from that package.

First you add the package to your library using ```library()``` then add the data to your R session by using ```data()```.

```{r}
library(wooldridge)
data('alcohol')
```

Now we have data on beer (the ```beer``` dataframe from above) and data from the ```alcohol``` dataframe available to use!


# Data Types and Structures in R

Once you've read your data into R, it's important to know how it's stored.

## Objects in R
Everything in R is an object. It has six basic data types (character, numeric, integer, logical, complex), and then each of these data types can be combined to form data structures, like lists and data frames.

To see what type of object your dataset is stored as, you can use "class()" or "typeof()". "Class()" identifies what type an object is from R's object-oriented-programming point of view. It determines how generic functions operate with it, and it's not a mutually exclusive classification. "Typeof()" the type of an R object. 


```{r}
class(4)
class('econometrics')
class(beer)
typeof(beer)
```

So why is the class "data.frame" and the type "list"? Data.frame and data.table are both collections (lists) of items, each item of the same length (so R considers each column as an item in the list). 

### Lists

Lists are the R objects which contain elements of different types like numbers, strings, vectors and another list inside it. A list can also contain a matrix or a function as its elements.

Basically, a list is a generic vector that contains other objects.

Lists are the most flexible data structure in R. They are a collection of elements without any restriction on the class, length, or structure of each element.

We can make a list of names:

```{r}
names <- list(c('Monet', 'Renoir', 'Degas'))
names
```

We can make a list of matrices:

```{r}
matrices <- list(matrix(c(1, 2, 3, 4, 5, 6), nrow = 2), matrix(c(7, 8, 9, 10, 11, 12), nrow = 2))
matrices
```

We can create a list with mixed data types:

```{r}
mixed_list <- list(c("Jan","Feb","Mar"), matrix(c(1,2,3,4,-1,9), nrow = 2),list("Red",12.3))
mixed_list
```
### Vectors

* You can create vectors, which can consist of multiple types (numeric, character, or logical), but all elements must be the same type. 

* To create a vector, you use c() for concatenate.

* You can get a subset of a vector: use my_vector[i] to get the ith element. You can also index a data frame.
  
  
```{r}
ages <- c(27, 28, 29, 30, 31)
ages[2]


ages[2:4]
```

* You can add, subtract, multiply, and divide vectors with numeric data.

```{r}
ages_sq <- ages^2
ages_sq

ages2 <- c(1, 3, 4, 6, 8)
age_diff = ages - ages2
age_diff

```

* You can do calculations with vectors: max(my_vector), min(my_vector), range(my_vector), length(my_vector), sum(my_vector), mean(my_vector), sort(my_vector), prod(my_vector) for the product of the elements, sd(my_vector) for the standard deviation, var(my_vector) for the variance.

```{r}
min(ages)
mean(ages)
prod(ages)
sd(ages)
```

### Data frames

Data frames are lists with restrictions. 

* You can't use the same name for two different variables. This is a good thing--you can give two elements the same name in lists, which can be confusing. R won't let you know that you've done it!
* All elements of a data frame are vectors.
* All elements of a data frame **have an equal length.** We will often have null values in our data; we will need to pay attention here. But we'll get into that more later!

The nice thing about data frames if you can select and do operations on rows, something you can't do with lists. You should use data frames whenever a column coincides with a variable and a row coincides with a single observation.

### More Data Structures

There are many more data structures in R, from vectors to matrices! [Here is an overview.](https://swcarpentry.github.io/r-novice-inflammation/13-supp-data-structures/)

# R Basics

In R, as in any coding language, certain symbols or combinations of symbols correspond to specific actions.

## Adding Notes to Your Code

* On any one line of code, anything after a "#" won't be executed. This is great for adding comments and notes to your code, or for preventing execution of lines without having to delete them.

```{r}
# This code won't be executed
# 2 + 2

# This code will be executed
2 + 2
```

## Arithmetic Operations

* You can do basic arithmetic operations, like addition, subtraction, multiplication, division, and exponentiation.

```{r}
# The expression below returns the value 0
4 - 4

# The expression below returns the value 8
4 + 4

# The expression below returns the value 16
4 * 4

# The expression below returns the value 1
4 / 4

# The expression below returns the value 256
4 ^ 4
```

* You can also do other functions, like logarithms, absolute values, or square roots.

```{r}
log(4) # This returns the natural log

exp(4)

abs(-4)

sqrt(4)
```

## Assigning Values to Variable Names

* You can assign numeric, character, and or logical values to variables.

```{r}
# Numeric variable
age <- 28 

# Character variable
name <- 'Fran' 

# Logical variable
# Is Fran a pilot?: (yes/no) <==> (TRUE/ FALSE)
is_pilot <- TRUE
```

## Functions on Vectors

* Parentheses are used to call functions. 

Basically everything in R is either an object or a function.

A function is a procedure that typically takes one or more objects as inputs, does something with those objects, then returns a new object. 

For example, below we will run the ```mean()``` function on the ```ages``` vector we created earlier. This function takes that vector object of numeric data, calculates the arithmetic mean of the numeric data, and then returns a scalar (a single number) as a result.

So when we used the ```ages``` vector and ran code like ```min(ages)```, we were calling a function. R has a ton of built-in functions to take use!  

Though it's archived, I really like [this overview](https://hbctraining.github.io/Intro-to-R/lessons/03_introR-functions-and-arguments.html) of functions in R.

```{r}
sum(ages)
mean(ages)
```

## Custom Functions

A function is a procedure that typically takes one or more objects as arguments (aka, inputs), does something with those objects, then returns a new object. 

For example, the ```mean()``` function we used above takes a vector object, like ages, of numeric data as an argument, calculates the arithmetic mean of those data, then returns a single number (a scalar) as a result. Squaring and adding vectors are other examples of functions.

But what about custom functions? These are very useful! While they may seem more complicated, just keep what you're trying to do in mind as you write them.

[For more detail on custom functions, see this chapter in YaRrr!](https://bookdown.org/ndphillips/YaRrr/functions.html). 

From YaRrr!:

Your custom functions will have the following 4 attributes:

1. Name: What is the name of your function? You can give it any valid object name. However, be careful not to use names of existing functions or R might get confused.

2. Arguments: What are the inputs to the function? Does it need a vector of numeric data? Or some text? You can specify as many inputs as you want.

3. Actions: What do you want the function to do with the inputs? Create a plot? Calculate a statistic? Run a regression analysis? This is where you’ll write all the real R code behind the function.

4. Output: What do you want the code to return when it’s finished with the actions? Should it return a scalar statistic? A vector of data? A dataframe?

Here’s how your function will look in R. When creating functions, you’ll use two new functions (Yes, you use functions to create functions! Very Inception-y), called function() and return(). You’ll put the function inputs as arguments to the function() function, and the output(s) as argument(s) to the return() function.

```
# The basic structure of a function
NAME <- function(ARGUMENTS) {

  ACTIONS

  return(OUTPUT)

}
```

Let's create a simple function using our ```ages``` vector.

```{r}
age_in_days_func <- function(age) {
  months <- age * 365
  return(months)
}
```

Let's give the function an age in years, and let it give us back the age in (approximate) days (ignoring leap years for simplicity):

```{r}
age_in_days_func(63)
```

But we had that whole list of ages--what if we wanted to get the ages in days for all of them?

# Loops

Again, big thank you to YaRrr! [Check out this chapter for more information.](https://bookdown.org/ndphillips/YaRrr/loops.html)

A loop is code that tells a program like R to repeat a certain chunk of code several times with different values of a loop object that changes for every run of the loop. In R, the format of a for-loop is as follows:

```
# General structure of a loop
for(loop.object in loop.vector) {

  LOOP.CODE

  }

```

As you can see, there are three key aspects of loops: The loop object, the loop vector, and the loop code:

1. Loop object: The object that will change for each iteration of the loop. This is usually a letter like i, or an object with subscript like column.i or participant.i. You can use any object name that you want for the index. 

2. Loop vector: A vector specifying all values that the loop object will take over the loop. You can specify the values any way you’d like (as long as it’s a vector). 

3. Loop code: The code that will be executed for all values in the loop vector. You can write any R code you’d like in the loop code - from plotting to analyses. R will run this code for all possible values of the loop object specified in the loop vector.

Let's write a loop to give us the age in months for everyone in our ages vector, and put all of those new ages in a new vector. We will  use the function we defined above inside of our loop.

```{r}
# create an empty vector
ages_in_days <- c()

for (i in ages) {
  age_in_days <- age_in_days_func(i)
  ages_in_days <- c(ages_in_days, age_in_days)
}

ages_in_days

```

# R Markdown

## Why use R Markdown?
R Markdown files are perfect for sharing. 

You can use them to both:

* save and execute code, and
* generate high quality reports that can be shared with an audience.

R Markdown was designed for easier reproducibility, since both the computing code and narratives are in the same document, and results are automatically generated from the source code. R Markdown supports dozens of static and dynamic/interactive output formats." (From [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/)--another fantastic resource!).

You will need to save your problem sets as either HTML or PDF documents for submission.

### Opening an R Markdown file

I usually do my work in R Markdown. To open a new R Script, go to File, New File, R Markdown.


### Markdown syntax

Markdown is a lightweight markup language that you can use to add formatting elements to plaintext text documents. 

You can use Markdown to make your documents look nice. You can add links, equations, or even images!

[Click here for a guide on formatting and math expressions in Markdown](https://bookdown.org/yihui/rmarkdown/markdown-syntax.html)


### Code chunks in R Markdown file

Within an R Markdown file, you will need to let R know which bits are writing and which bits are code to execute. To have a code chunk, you put:

" ```{r} "

INSERT CODE HERE

" ``` "

Only without the quotation marks, of course! Then R will execute whatever is in between the "``` {r}" and the closing "```".

### Including Plots in R Markdown files

You can also embed plots, for example. Just put plot() in your code chunk.

```{r, fig.alt = "A scatterplot of temperature vs. pressure"}
plot(pressure)
```


### Saving as HTML

If you have an R Markdown file open, you need to go to "Knit" (it's at the top of the window), and choose "Knit to HTML". That's it!

It will save to the same folder where you've saved your R file.

But be careful--if you knit a new HTML file without changing the name of your R file, it will overwrite what you've done before! If you want to keep track of different versions, you will need to change the name of your R file before knitting.

### Saving as PDF

Knitting to a PDF requires a little extra work. To save this R Markdown file as a PDF, you can use TinyTeX. It's a lightweight LaTeX distribution. 

If you don't have TinyTeX installed, you can use the first three lines of the following code to then save your R Markdown file as a PDF:

```{r}
#install.packages('tinytex')
#tinytex::install_tinytex()
#install.packages('rmarkdown')

#library(rmarkdown)
#library(tinytex)
```

Then once you're ready, go to "Knit" again. This time, choose "Knit to PDF." Your PDF will be saved in the same folder where your R file is saved.

For this class, I will provide sample notebooks in .html and .Rmd formats. But knitting to PDF using LaTeX is a great way to write your paper!


