---
title: 'Why $\hat{p}$ Uses $n$ Twice: Proportion as an Average & the 1/n Variance Rule'
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    highlight: tango
#    includes:
#      in_header: wcag-overrides.html

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Average vs. Variance: Is it double counting?

In a survey of 600 likely voters, 272 preferred the incumbent and 328 preferred the challenger. Let $\hat{p}$ be the fraction of all likely voters who preferred the incumbent at the time of the survey.

- Sample size: $n=600$
- Count for incumbent: $272$  
- Sample proportion: $\hat{p} = \frac{272}{600}$


We want the standard error of $\hat{p}$ using the usual plug-in estimator.

# Two Roles for $n$: Definition vs. Variability

## 1. $\hat{p}$ is a sample average
Define indicator variables $(Y_i\in {0,1}): 1$ if voter $i$ prefers the incumbent, 0 otherwise. Then

$\hat{p} = \frac{1}{n}\Sigma_{i=1}^n Y_i$

Here, the $n$ in the denominator is only used as a definition, it’s how we form an average.

## 2. Averages are $1/n$ as variable as single draws
Because $\hat{p}$ is the average of $n$ independent draws, its variance is reduced by a factor of $1/n$:

${Var}(\hat{p})$
$={Var}\left(\frac{1}{n}\Sigma_{i=1}^n Y_i\right)$
$=\frac{1}{n^2}\Sigma_{i=1}^n{Var}(Y_i)$

Since each $Y_i$ has variance $p(1-p)$:

$Var(\hat{p}) = \frac{1}{n^2} \cdot n \cdot p(1-p) = \frac{p(1-p)}{n}$.

When $n$ shows up again it has a statistical role, since it captures how when we average across more observations it makes the estimator $\hat{p}$ less noisy. 

A single Bernoulli draw is very noisy, since it can only take on the values of 0 or 1. But when we average across a lot of independent draws the ups and downs tend to cancel out and the more observations we have, the less variability in the average.

# Plug-in Estimator and Numerical SE

Since $p$ is unknown, we use $\hat{p}$ in its place:
$\widehat{\mathrm{Var}}(\hat p)=\frac{\hat p(1-\hat p)}{n}, \qquad \mathrm{SE}(\hat p)=\sqrt{\frac{\hat p(1-\hat p)}{n}}.$


```{r compute-se}
n <- 600
x <- 272
p_hat <- x / n
var_hat <- p_hat * (1 - p_hat) / n
se_hat <- sqrt(var_hat)

list(
  n = n,
  x = x,
  p_hat = p_hat,
  var_hat = var_hat,
  se_hat = se_hat
)
```

**Interpretation.** 

$\mathrm{SE}(\hat p)$ is the standard deviation of the sampling distribution of $\hat{p}$ (how much $\hat{p}$ would fluctuate across repeated samples of size $n$). It is not the population standard deviation of $Y$, which is $\sqrt{p(1-p)}$.

# Averaging dice rolls

Rolling one die is noisy but averaging 600 rolls is much calmer. When we average many independent outcomes it's about $1/n$ as variable (in variance) as a single outcome. 

Each $Y_i$ is like a “0/1 die” with mean $p$ and variance $p(1-p)$.  

The sample proportion $\hat p$ is the average, so its variance is $p(1-p)/n$, and its standard deviation (the standard error) is $\sqrt{p(1-p)/n}$.

## Quick Simulation 

Let's simulate many repeated samples under a Bernoulli model with probability equal to our observed $\hat p$.  
We can compare the empirical variance of $\hat p$ across repetitions to the plug-in estimator $\hat p(1-\hat p)/n$.

```{r}
# reminding ourselves of the values from before
list(
  n = n,
  x = x,
  p_hat = p_hat,
  var_hat = var_hat,
  se_hat = se_hat
)
```


```{r simulate}
set.seed(123)

p_sim <- p_hat          # use the observed proportion as the true p for simulation
n <- 600
R <- 20000              # repetitions

# simulate R sample proportions of size n
p_hats <- replicate(R, mean(rbinom(n, size = 1, prob = p_sim)))

emp_var <- var(p_hats)
theory_var <- p_sim * (1 - p_sim) / n

c(
  empirical_variance_of_p_hat = emp_var,
  theoretical_variance_of_p_hat = theory_var,
  ratio_empirical_to_theoretical = emp_var / theory_var
)

```

The empirical variance closely matches the theoretical $\hat p(1-\hat p)/n$, and the ratio hovers near 1.

## Key Takeaways

* The first $n$ in $\hat p = \frac{1}{n}\sum Y_i$ is definitional: a proportion is an average of 0 and 1’s.

* The second $n$ in $\mathrm{Var}(\hat p)=\frac{p(1-p)}{n}$ is statistical since averaging $n$ independent observations reduces variance by $1/n$.

* The plug-in estimator gives $\widehat{\mathrm{Var}}(\hat p)=\frac{\hat p(1-\hat p)}{n}, \quad \mathrm{SE}(\hat p)=\sqrt{\frac{\hat p(1-\hat p)}{n}}.$

* ${SE}(\hat{p})$ is the standard deviation of the sampling distribution of the estimator, not the population standard deviation of $Y$.

