---
title: "In-Class Exercises 1-21"
output: 
  html_document:
    toc: yes
    toc_depth: 2
    df_print: kable
    highlight: tango
#    includes:
#      in_header: wcag-overrides.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
```

# 1. Random Variables & Distributions

Warm‑up (no code)

1. In your own words, what is a random variable?


2. What does it mean to say a variable follows a normal distribution?



```{r}
# Draw 1,000 observations from a normal distribution

```


## Task 1: Compute the mean and standard deviation of your observations manually using mean() and sd().

* How close are they to the values we specified (50 and 10)?



```{r}
# calculate the mean and standard deviation
```


Discussion:

* Why aren’t the sample values exactly 50 and 10?



## Task 2: Visualizing the Distribution

```{r}
# create a histogram of your simulated data

```

Group discussion questions:

* Is the distribution symmetric?

* Where is most of the mass concentrated?


# 2. Probability & Events


Concept check

Suppose:

* $P(A) = 0.4$

* $P(B) = 0.3$

* $P(A \cap B) = 0.1$

* Compute $P(A \cup B)$.



* Are A and B independent? Why or why not?




## Task 3: Empirical Probabilities via Simulation

```{r}
# Roll a die 10,000 times and save the sample



# Calculate the probability of rolling a 6 from your sample

```


Next:

* Estimate the probability of rolling an even number from your sample.

* Estimate P(roll ≥ 4).

```{r}
# probability of rolling an even number

# probability P(roll >= 4)

```


Discussion:

* Why do these values get closer to the theoretical probabilities as sample size increases?



# 3. Law of Large Numbers

## Task 4: Convergence of Sample Means

Generate two samples, one of size 10 and one of size 5,000.

Each of these samples should draw from the 

```{r}
# Generate increasing sample sizes
n <- seq(10, 5000, by = 10)
means <- sapply(n, function(i) mean(rnorm(i, 50, 10)))



plot(n, means, type = "l",
main = "Law of Large Numbers",
xlab = "Sample size",
ylab = "Sample mean")
abline(h = 50, col = "red")
```


Questions

1. What happens to the sample mean as n grows?



2. Why does this matter for empirical research?



# 4. Central Limit Theorem

## Task 5: Sampling Distribution of the Mean

```{r}
# Draw 1000 sample means from a skewed distribution
sample_means <- replicate(1000, mean(rexp(40)))

# create a histogram of the sample means

```


Group discussion

* Is the original distribution normal?


* Is the sampling distribution approximately normal?


* Why is this result important for inference?



# 5. Descriptive Statistics

## Task 6: Working with Real Data

```{r}
# add the mtcars data to your R session

```

Next:

* Compute mean, median, and standard deviation of `mpg`.

* Interpret each statistic.

```{r}
# mean

# median

# sd
```



## Task 7: Group Comparisons

```{r}
# Compare the mean mpg for automatic vs manual transmission

```

Questions:

* Which group has higher fuel efficiency?



* Is this a causal claim? Why or why not?


# 6. Covariance & Correlation

## Task 8: Correlation in Practice

```{r}
# calculate the correlation between horsepower (hp) and mpg

```

Questions:

* Is this relationship positive or negative?


* How strong is it using the correlation guidelines?



## Task 9: Visualizing Correlation

```{r}
# create a scatterplot with horsepower on the x-axis and mpg on the y-axis
# draw a regression line through the scatterplot

```

Discussion:

* Does the plot match the correlation sign?



* What patterns do you see?


# 7. Reflection & Wrap‑Up


* **What did simulation help you understand today?**


* **Which concept (LLN, CLT, correlation, etc.) still feels fuzzy?**



* **How could these tools apply to your own research questions?**






