---
title: "Topic 3 In-Class Notebook"
output: 
  html_document:
    toc: yes
    toc_depth: 2
    df_print: kable
    highlight: tango
#    includes:
#      in_header: wcag-overrides.html

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(wooldridge)
#install.packages('sandwich')
library(sandwich)
library(lmtest)
#install.packages('kableExtra')
library(kableExtra)

data("pntsprd")
```

## Dataset + variables

We will use `pntsprd` from **wooldridge** (NCAA men's basketball games).

Key variables:

* `scrdiff`: actual score difference (favored team score − underdog score)
* `spread`: Las Vegas point spread (favored team by how many points?)
* `favhome`: 1 if the favored team is at home, 0 otherwise

(If you want to see all variables: `?pntsprd`.)

```{r}
?pntsprd
```


---

## Problem 1 — Model, notation, and interpretation (warm-up, pntsprd)

We’ll study how well the Las Vegas point spread predicts the actual score difference.

### 1a. Fit the SLR model in R

```{r}
mod1 <- lm(scrdiff ~ spread, data = pntsprd)
summary(mod1)
```

```{r}
source('reporting.R')
```

```{r}
mk_coeftab(mod1, #model
           mode = 'simple', #modes are simple and full
           term_labels = c(
             spread = "Las Vegas Point Spread"
           ),
           caption = "Model: Score Difference on Point Spread")
```


### 1b. Write the population model and the sample (estimated) model

In math notation:

- Population model:


$scrdiff_i = \beta_0 + \beta_1 spread_i + u_i$

- Estimated (sample) model:


$\widehat{scrdiff}_i = \hat\beta_0 + \hat\beta_1 spread_i$


### 1c. Interpret the slope in context

In one sentence, interpret your estimated slope $\hat\beta_1$ from `summary(mod1)`.

The value of $\hat\beta_1 = 1.002$

A one point increase in the Vegas spread (the x variable) is associated with an average increase of $\hat\beta_1$ (1.002) points in the actual score difference.

### 1d. What does the error term contain?

In one sentence: what kinds of things are in $u_i$ here?

$u_i$ includes everything affecting game score differences besides the spread: playing at home, semi finals/ finals, injuries, travel, measurement noise, weather.

---

## Problem 2 — OLS by hand 

Given 5 observations:

| i | X | Y |
|---|---:|---:|
|1|1|2|
|2|2|3|
|3|3|5|
|4|4|4|
|5|5|6|

1) Compute $\bar X$, $\bar Y$.  

2) Compute the OLS slope $\hat\beta_1 = \frac{\sum (X_i-\bar X)(Y_i-\bar Y)}{\sum (X_i-\bar X)^2}$. 

3) Compute $\hat\beta_0 = \bar Y - \hat\beta_1 \bar X$.

```{r}
# put your work in this code chunk
x <- c(1, 2, 3, 4, 5)
y <- c(2, 3, 5, 4, 6)

xbar <- mean(x)
ybar <- mean(y)

b1 <- sum((x - xbar)*(y - ybar))/sum((x - xbar)^2)
b0 <- ybar - b1 * xbar

b1
b0

coef(lm(y ~ x))
```


---

## Problem 3 — Fitted values, residuals, and “fit” measures 

Using your fitted line from Problem 2:

1) Compute $\hat Y_i$ and residuals $\hat u_i = Y_i - \hat Y_i$ for all 5 points.  

$\hat y_i = 1.3 + 0.9 x$

2) Compute $SSE =\sum \hat u_i^2$. 
Sum of squared errors. Proportion of variability that the model does not explain.

$\hat u_i = y_i - \hat y_i$


3) Compute $R^2 = SSR/TSS = 1 - SSE/TSS$
TSS (SST): represents the variation in y around its mean.

$TSS = \Sigma^n_{i=1} = (y_i - \bar y)^2$

SSR: Regression sum of squares, the variability that the model explains compared to a model with no variables.

$SSR = \Sigma^n_{i=1}(\hat y_i - \bar y)^2$

R squared is the proportion of variation in y explained by the regression.

NOTE: be consistent between Residual / Error sum of squares!

.  
4) Compute $SER = \sqrt{SSE/(n-2)}$.

```{r}
# put your work in this code chunk
yhat <- b0 + b1 * x
uhat <- y - yhat
SSE <- sum(uhat^2)
TSS <- sum((y - ybar)^2)
SSR <- sum((yhat - ybar)^2)

rsq <- SSR / TSS  
rsq

rsq_2 <- 1 - SSE / TSS
rsq_2

SER <- sqrt(SSE / (length(y) - 2))
SER

test_mod <- lm(y ~ x)

c(SSE = sum(resid(test_mod)^2), 
  TSS = sum((y - ybar)^2), 
  R2 = summary(test_mod)$r.squared, 
  SER = sqrt(sum(resid(test_mod)^2) / (length(y) - 2))
  )
```

---

## Problem 4 — The big 4 SLR assumptions (pntsprd)

Considering the model

$scrdiff_i = \beta_0 + \beta_1 spread_i + u_i$


State each assumption in a sentence and give one example where it might fail.

1) Linear in parameters

2) i.i.d. sampling 

3) Sampling variation in the explanatory variables

4) $E(u_i\mid spread_i)=0$



---

## Problem 5 — Reading regression output / testing a specific null hypothesis

Using your `mod1` estimates, test the null $H_0:\beta_1 = 1$

Compute the t-stat for the null that the slope equals 1.

You can do this by hand using the estimate and standard error from `summary(mod1)`:


$t = \frac{\hat\beta_1 - 1}{SE(\hat\beta_1)}$

```{r}
# compute your t-stat here
mod1 <- lm(scrdiff ~ spread, data = pntsprd)
coef(mod1)
b1 <- coef(mod1)["spread"]
se1 <- sqrt(vcov(mod1)["spread", "spread"])

t_stat <- (b1 - 1) / se1

alpha <- 0.05

df <- mod1$df.residual
nrow(pntsprd) - 2

# 5 % two sided critical value
t_crit <- qt(1 - alpha/2, df)

# if abs(t_stat) > t_crit, then reject the null
abs(t_stat) > t_crit

?qt

```


You can also use `linearHypothesis`, but we'll stick to the "by-hand" method for today.

---

## Problem 6 — Confidence intervals / test equivalence

### 6a. 95% confidence interval for $\beta_1$

```{r}
confint(mod1)
```

Does the CI include 1? Use that to decide whether you’d reject $H_0:\beta_1=1$ at 5%.

Since the confidence interval includes 1, we fail to reject the null hypothesis. We can reject the null that $H_0:\beta_1=0$.

---

## Problem 7 — Dummy regressor = difference in means 

Fit:

```{r}
mod2 <- lm(scrdiff ~ favhome, data = pntsprd)
summary(mod2)
```

1) What is the predicted mean `scrdiff` when `favhome=0`? When `favhome=1`? 

2) What is the difference in means (home − not home)?



---

## Problem 8 — Robust standard errors (sandwich + lmtest)

Compare usual vs robust standard errors for `mod1`

```{r}
# Usual (homoskedastic) SEs:
coeftest(mod1)

# Robust (HC1) SEs:
coeftest(mod1, vcov. = vcovHC(mod1, type = "HC1"))
```

1) Do the *coefficients* change?  

2) Do the *standard errors* change?  

3) What happens to the t-stat / p-value for `spread`?

4) If robust SEs are different, how would that change your CI intuition?




