---
title: "In-Class Activity: Limited Dependent Variable Models with the `charity`/`gift` Data"
author: "ECON 8300"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Packages + data

Today we will use the **`charity`** data from the **`wooldridge`** package. The outcome of interest is **`gift`**, the amount donated. 

We will also use **`respond`**, which is already a binary indicator for whether a gift was made.

```{r}
need <- c("tidyverse", "wooldridge", "sandwich", "lmtest", "margins", "AER", "kableExtra")
have <- need %in% rownames(installed.packages())
if (any(!have)) install.packages(need[!have])

invisible(lapply(need, library, character.only = TRUE))

df <- charity %>%
  as_tibble() %>%
  mutate(
    respond = as.integer(respond),
    gift = if_else(is.na(gift), 0, gift)
  )
```

# The data

## Activity 1: Explore the distribution of `gift`

First take a look at the dataset:

```{r}
?charity
```


Create a quick overview of the data.

### Your tasks

1. Display the dimensions of the data.
2. Produce summary statistics for:
   - `gift`
   - `respond`
   - `mailsyear`
   - `giftlast`
   - `avggift`
   - `propresp`
3. Compute the share of observations with `gift == 0`.
4. Create a histogram of `gift`.

### Answer

```{r}
# Activity 1
```

### Write-up

In 2–3 sentences: what makes `gift` a good example of a limited dependent variable?

**Answer:**



## Activity 2: Compare donors and non-donors

Create grouped summary statistics by `respond`.

### Your tasks

1. Group by `respond`.
2. Report means for:
   - `gift`
   - `mailsyear`
   - `giftlast`
   - `avggift`
   - `propresp`
   - `weekslast`
3. Write 2 sentences comparing the groups.

### Answer

```{r}
# Activity 2
```

### Write-up

**Answer:**




# Linear Probability Model

## Activity 3: Estimate an LPM for response

We now model whether someone gives at all.

Estimate the model


$respond_i = \beta_0 + \beta_1 mailsyear_i + \beta_2 giftlast_i + \beta_3 avggift_i + \beta_4 propresp_i + \beta_5 weekslast_i + \beta_6 resplast_i + u_i$

### Your tasks

1. Estimate the model by OLS.
2. Report coefficients with **HC1 robust SEs**.
3. Store fitted values.
4. Check whether any fitted values are below 0 or above 1.

### Answer

```{r}
# Activity 3
```

### Write-up

Why can fitted values outside $[0,1]$ be a problem in an LPM?

**Answer:**





# Binary response models

## Activity 4: Estimate logit and probit

Estimate logit and probit versions of the same response model.

### Your tasks

1. Estimate a **logit** model.
2. Estimate a **probit** model.
3. Report coefficients with **HC1 robust SEs**.
4. Compare the signs of the coefficients across LPM, logit, and probit.

### Answer

```{r}
# Activity 4
```

### Write-up

Do the signs line up across the three models (including LPM)? Which variables appear most important?

**Answer:**



## Activity 5: Compute average marginal effects

For both logit and probit, compute **average marginal effects** for:

- `avggift`
- `mailsyear`

### Your tasks

1. Use `margins()` or compute the AMEs manually.
2. Report the AMEs.
3. Interpret one AME in a sentence.

### Answer

```{r}
# Activity 5
```

### Write-up

Interpret the AME for `mailsyear` in probability terms.

**Answer:**




## Activity 6: Compare predicted probability curves

Create one figure with **three curves** showing predicted probability of response as a function of `avggift`:

- LPM
- logit
- probit

Hold the other regressors at their sample means.

### Your tasks

1. Reuse or rebuild a prediction grid.
2. Add predictions from all three models.
3. Plot all three curves on one graph with a legend.

### Answer

```{r}
# Activity 6
```

### Write-up

What is the biggest practical difference between the LPM curve and the logit/probit curves?

**Answer:**



# Tobit

## Activity 7: Model the amount donated with Tobit

Now switch from the binary outcome to the donation amount itself.

Estimate a Tobit model for


$gift_i = \max(0, gift_i^*)$

with latent equation


$gift_i^* = \beta_0 + \beta_1 mailsyear_i + \beta_2 giftlast_i + \beta_3 avggift_i + \beta_4 propresp_i + \beta_5 weekslast_i + \beta_6 resplast_i + u_i$

### Your tasks

1. Estimate a Tobit model with left-censoring at 0.
2. Estimate OLS for `gift` using the same regressors.
3. Compare the Tobit and OLS coefficients.
4. Write 3–4 sentences explaining:
   - why `gift` is a corner-solution outcome,
   - why Tobit may be more appropriate than OLS here,
   - why a Tobit coefficient is **not** automatically the marginal effect on observed `gift`.

### Answer

```{r}
# Activity 7
```

### Write-up

**Answer:**



# Wrap-up

## Exit question

Write one short paragraph comparing these four models:

- LPM
- logit
- probit
- Tobit

Your paragraph should address:

1. what each model is trying to explain,
2. one advantage of LPM,
3. one reason logit/probit are often preferred for binary outcomes,
4. one reason Tobit is useful for `gift`.

**Answer:**



# Optional extension

If you finish early, try one of these:

1. Replace `avggift` with `giftlast` on the prediction graph and compare.
2. Compute the proportion of observations predicted outside $[0,1]$ by the LPM.
3. For the Tobit model, compare predicted values for two hypothetical people:
   - low prior giving / low response history
   - high prior giving / high response history
