---
title: "In-Class Activity: Simple Linear Regression"
output: 
  html_document:
    toc: yes
    toc_depth: 2
    df_print: kable
    highlight: tango
    includes:
      in_header: wcag-overrides.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Learning goals

By the end of this activity, you should be able to:

- describe the population and sample simple regression models;
- estimate an OLS regression in R;
- calculate the OLS slope and intercept from their formulas;
- interpret coefficients, fitted values, residuals, and $R^2$; and
- distinguish an association from a causal effect.


# 1. Set up and meet the data

We will use `wage2` from the `wooldridge` package. Each observation represents one person. Our main variables are:

- `wage`: monthly earnings in dollars;
- `lwage`: the natural log of monthly earnings; and
- `educ`: years of education.

If needed, install a package once in the Console with `install.packages("package_name")`. Do not include installation commands in the notebook.

```{r packages}
library(wooldridge)
library(tidyverse)
library(knitr)
```

Use the help file, `glimpse()`, and `summary()` to explore the data.

(`glimpse()` is from `tidyverse`.)

```{r inspect-data}
# View the documentation for wage2

# Inspect the structure of wage2

# Summarize educ, wage, and lwage

```

1. What is the unit of observation?



2. How many observations are in the data? Are any values missing for `educ`, `wage`, or `lwage`?



3. Record the mean years of education and mean monthly earnings. Why is it useful to inspect descriptive statistics before fitting a model?



# 2. Visualize the relationship

Create a scatterplot with `educ` on the horizontal axis and `lwage` on the vertical axis. Add a fitted linear regression line, informative labels, and alt text. 

```{r scatterplot, fig.alt="Scatterplot of log monthly earnings against years of education, with a fitted linear regression line."}
# Create the scatterplot here

```

4. Describe the direction, strength, and form of the relationship. Identify anything else that stands out.



5. Calculate the correlation between `educ` and `lwage`. Does it agree with what you saw in the plot? Explain.

```{r correlation}
# Calculate the correlation here

```



# 3. Connect the population model to the sample

The population regression model is

$$
Y_i = \beta_0 + \beta_1X_i + u_i,
$$

where $u_i$ includes all factors other than $X_i$ that affect $Y_i$.

For this application, write the population model using the variable names `lwage` and `educ`.



6. In words, what does $\beta_1$ represent?



7. Name two factors that could belong in $u_i$. For each, explain why it might also be related to education.



The sample regression line is

$$
\widehat{Y}_i = \widehat{\beta}_0 + \widehat{\beta}_1X_i.
$$

The residual for observation $i$ is

$$
\widehat{u}_i = Y_i-\widehat{Y}_i.
$$

8. What does a positive residual mean here? What does a negative residual mean?



# 4. Calculate OLS estimates from the formulas

OLS chooses the intercept and slope that minimize the sum of squared residuals:

$$
\min_{b_0,b_1}\sum_{i=1}^{n}(Y_i-b_0-b_1X_i)^2.
$$

For simple linear regression,

$$
\widehat{\beta}_1 =
\frac{\sum_{i=1}^{n}(X_i-\bar{X})(Y_i-\bar{Y})}
{\sum_{i=1}^{n}(X_i-\bar{X})^2}
$$

and

$$
\widehat{\beta}_0=\bar{Y}-\widehat{\beta}_1\bar{X}.
$$

Complete the code to calculate both coefficients. Do not use `lm()` yet.

```{r coefficients-by-hand}

x_bar <- # mean of educ
y_bar <- # mean of lwage

numerator <- # sum of (educ - x_bar) * (lwage - y_bar)
denominator <- # sum of (educ - x_bar)^2

beta_1_hat <- # numerator divided by denominator
beta_0_hat <- # intercept formula

beta_0_hat
beta_1_hat
```

9. Why do we square the residuals instead of just adding them?



10. Check two properties of an OLS regression with an intercept:

- the residuals sum to approximately zero and
- the sample regression line passes through $(\bar{X},\bar{Y})$.

```{r verify-properties}
# Calculate fitted values and residuals using your hand-calculated coefficients

# Check the sum of the residuals

# Compare y_bar with the fitted value at x_bar

```

**What did you find?**



# 5. Estimate the model with `lm()`

Fit the model

$$
lwage_i=\beta_0+\beta_1 educ_i+u_i.
$$

```{r fit-model}
# Fit the model and save it as wage_model

# Display the model summary

```

11. Check that the `lm()` coefficients match your calculations. Write the estimated regression equation.



12. Interpret the slope using both approximations below.

- Approximate: a one-unit increase in $X$ is associated with a $100\widehat{\beta}_1\%$ change in $Y$.
- Exact: $100(e^{\widehat{\beta}_1}-1)\%$.

```{r slope-percent}
# Calculate the approximate percent change

# Calculate the exact percent change

```


13. Interpret the intercept. Is it actually useful here? Use the range of `educ` to support your answer.


```{r education-range}
# Find the minimum and maximum of educ

```



# 6. Predictions and residuals

Use `predict()` to estimate `lwage` for people with 12 and 16 years of education. Then convert each prediction back to dollars with `exp()`.

```{r predictions}
new_people <- tibble(educ = c(12, 16))

# Add predicted lwage and predicted wage to new_people

```

14. Report and compare the predicted monthly earnings. Why is the dollar difference not constant even though the regression is linear in `lwage`?



Add fitted values and residuals to a copy of the data. Then identify the observations with the largest positive and largest negative residuals.

```{r residual-observations}
wage2_results <- wage2 |>
  mutate(
    fitted_lwage = # fitted values,
    residual = # residuals 
  )
  

# Show the observations with the largest positive and negative residuals

```

15. Interpret each of those two residuals in context.



# 7. Measure model fit

Using the standard sums-of-squares notation:

$$
TSS=\sum(Y_i-\bar{Y})^2, \qquad
SSR=\sum(Y_i-\widehat{Y}_i)^2 = \sum\hat{u_i}^2, \qquad
ESS=\sum(\widehat{Y}_i-\bar{Y})^2.
$$

They satisfy $SST=SSR+SSE$, and

$$
R^2=\frac{ESS}{TSS}=1-\frac{SSR}{TSS}.
$$

Calculate the three sums of squares and $R^2$ from the fitted values and residuals. Compare your result with the value from `summary(wage_model)`.

```{r r-squared-by-hand}
TSS <- # total sum of squares
SSR <- # sum of squares residuals
ESS <- # explained sum of squares

r_squared_hand <- # calculate R-squared

# Verify the decomposition and compare R-squared values

```

16. Interpret $R^2$ in context. Does a small $R^2$ mean the slope estimate is useless? Explain.


# 8. Examine residuals and causal language

Create a residual plot with `educ` on the horizontal axis and the residuals on the vertical axis. Include a horizontal line at zero and alt text.

```{r residual-plot, fig.alt="Residuals from the regression of log monthly earnings on education plotted against years of education, with a horizontal reference line at zero."}

# create the residual plot here
```

17. Are the residuals randomly scattered around zero, or do you see a systematic pattern? What does this plot reveal—and what can it not prove—about the zero conditional mean assumption $E(u\mid X)=0$?


18. A classmate writes: “One additional year of education causes earnings to rise by about 6%.” Rewrite this conclusion so it is supported by this regression. Then explain what extra evidence or research design would be needed to make a causal claim.



# 9. Final synthesis

In 3–5 sentences, summarize what this analysis shows. Include the direction and magnitude of the estimated relationship, model fit, and one important limitation.



# Optional extension

Fit a level-level model using `wage` instead of `lwage`. Compare its plot, coefficient interpretation, residual plot, and $R^2$ with the log-level model. Which specification seems more useful, and why?

```{r extension, eval=FALSE}
# Optional extension code

```

