---
title: "In-Class Activity: Further Inference in Multiple Regression"
output:
  html_document:
    toc: yes
    toc_depth: 2
    df_print: kable
    highlight: tango
    includes:
      in_header: wcag-overrides.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
#library(here)
#here::i_am("in-class-notebooks/in-class-activity-multiple-regression-inference.Rmd")
```

# Plan for today


## Goals
Practice the core inference ideas in multiple regression:

* single-coefficient tests & confidence intervals,
* heteroskedasticity-robust standard errors,
* joint hypothesis testing (F-tests),
* prediction / prediction intervals,
* quick diagnostics (RESET, VIF).


## Packages + data

We’ll use the **gpa2** data (Wooldridge). If `wooldridge` is not installed, install it.

```{r}
need <- c("tidyverse", "wooldridge", "sandwich", "lmtest", "car", "kableExtra")
have <- need %in% rownames(installed.packages())
if (any(!have)) install.packages(need[!have])
invisible(lapply(need, library, character.only = TRUE))
#source(here("scripts", "reporting.R"))
source("reporting.R")

data("gpa2")

```

Quick variable recap:

* `colgpa`: college GPA (dependent variable)
* `tothrs`: total study hours
* `athlete`: 1 if student-athlete
* `sat`: SAT score

We’ll estimate:


$colgpa = \beta_0 + \beta_1 tothrs + \beta_2 athlete + \beta_3 sat + u$

---

# Task 1 — Fit the model + read the output 

1. Estimate the model above and store it as `mod1`.
2. Print the regression summary.
3. In **1–2 sentences**, interpret **the slope on `tothrs`** in plain English.

```{r, results = "asis"}
# fit the model
mod1 <- 

# view results

```

**Interpretation (write here):**  


---

# Task 2 — Test a single coefficient + build a 95% CI 

## 2A. Hypothesis test (two-sided)
Test:


$H_0:\beta_{tothrs}=0 \quad \text{vs} \quad H_1:\beta_{tothrs}\neq 0$

1. Extract the estimate and the usual (homoskedastic) standard error for `tothrs` and `sat`
2. Compute the **t statistic** by hand for `tothrs` and `sat`
3. Compute the **two-sided p-value** using `pt()` for both variables.

```{r}
# extract beta_hat and se_hat for tothrs
b1  <- 
se1 <- 

b2 <- 
se2 <- 

#  t statistic for H0: beta1 = 0
t_stat1 <- 
t_stat2 <- 

#  two-sided p-value
p_val1 <- 
p_val2 <- 

```

## 2B. 95% confidence interval

Construct a 95% CI for $\beta_{tothrs}$ and $\beta_{sat}$ using:

$\hat\beta \pm t_{0.975, df}\cdot SE(\hat\beta)$

```{r}
# critical value
t_crit <- 

# CI endpoints
ci1 <- 


ci2 <- 


```

**Interpret the confidence intervals:**



---

# Task 3 — Robust standard errors (HC1) (Driver 2)

Real data often have heteroskedasticity. Re-do inference for `tothrs` using **HC1 robust** SEs.

1. Compute a robust variance-covariance matrix with `vcovHC(mod1, type="HC1")`.
2. Use `coeftest()` to print the coefficient table with robust SEs.
3. Compare the usual vs robust SE for `tothrs` (which is bigger? Why?).

```{r}
res_df <- tibble(
  fitted_m1 = fitted(mod1),
  resid_m1 = resid(mod1)
)
```


```{r, fig.alt = "residuals plot for model 1"}
# create a residual / fitted values plot for model 1 

```



```{r}
# get the robust VCOV matrix
V_hc1 <- 

# get the robust coefficient test table for model 1 using coeftest()
coeftest(mod1, vcov. = V_hc1)

# compare SEs for tothrs (robust and usual)
se1_robust <- 
```

**1–2 sentence takeaway about residuals plots and robust standard errors:**  


---

# Task 4 — Joint hypothesis test (F-test) (Driver 2)

We want to test whether `tothrs` and `athlete` matter *jointly*:


$H_0:\beta_{tothrs}=0 \ \text{and}\ \beta_{athlete}=0 \ \text{and} \beta_{female} = 0 \quad \text{vs} \quad H_1: \text{at least one is nonzero}.$

## 4A. Do it “by hand” using restricted vs unrestricted SSE

Steps:
1. Fit a **restricted** model that drops `tothrs` and `athlete` and `female` (keep `sat`).
2. Extract SSE (sum of squared residuals) for restricted and unrestricted.
3. Compute the F statistic:


$F = \frac{(SSE_R - SSE_U)/q}{SSE_U/(N-p)}$

where $q$ is the number of restrictions $p$ is number of parameters in the unrestricted model (including intercept).

```{r}
# restricted model (only sat)

# SSEs


# pieces for F


# F stat and p-value

```

## 4B. Do it with `linearHypothesis()`

```{r}
linearHypothesis()
```

**F-stat interpretation (3-4 sentences):**  



---

# Task 5 — Prediction + prediction interval (Driver 1)

Let’s predict GPA for a hypothetical student:
- `tothrs = 20`
- `athlete = 0`
- `sat = 1100`

1. Create a one-row data frame called `new_student`.
2. Compute the predicted mean GPA.
3. Compute a **95% prediction interval**.

```{r}
# define new observation


# point prediction


# 95% prediction interval

# 95% confidence interval (compare)
```

**Question:** Why is the *prediction interval* wider than a confidence interval for the conditional mean?  
*(Answer in a few sentences.)*  



---

# Task 6 — Quick diagnostics: RESET + VIF (Driver 2)

## 6A. RESET (functional form)

Use `resettest()` (from `lmtest`) to check whether adding powers of fitted values suggests misspecification. First run a RESET test:

```{r}
# RESET test

```

**Interpretation:**  
What do your results tell you about your model's specification? What is the RESET test doing "under the hood"?


## 6B. Collinearity via VIF
Compute variance inflation factors (VIF). 

```{r}
#  VIFs

```

**Interpretation:**  
Why is it called a "Variance Inflation Factor"? What does the presence of multicollinearity do to the coefficients? What do your results tell you about your model?





---

# Wrap-up (2 minutes)

1. One thing you feel solid on:
2. One thing you want to review before the homework:
