---
title: "In-Class Activity: Exploring Data with R"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Purpose

In this activity, you will work with a partner to explore the built-in `mtcars`
data. The goal is not to memorize R commands. The goal is to practice reading
code, writing small pieces of code, interpreting output, and solving problems
together.


## Pair-programming roles

- **Driver:** controls the keyboard and writes the code.
- **Navigator:** reads the instructions, checks the code, predicts results, and
  searches the notebook or R help when the pair gets stuck.

Both partners should explain what the code is doing. Switch roles whenever the notebook tells you to switch. If one person has more R experience, that person should begin as the navigator.

**Pair members:**  
Driver (starting): 
Navigator (starting): 

# Part 1: Meet the data (10--15 minutes)

The `mtcars` data frame is included with R. Each row represents a car model and each column contains a characteristic of that car.

```{r load-data}
cars <- mtcars
```

```{r}
x <- 15:20
x
y <- 10

z <- x*y
z
```

## Task 1.1: Inspect the object

Before running each command, the navigator should predict what kind of information it will return. Then run the code and briefly describe the result.

```{r inspect-data}
class(cars)
typeof(cars)
dim(cars)
names(cars)
cars
```
Other data types: logical, integer (whole numbers), double (decimal numbers), characters, 
1. What is the class of `cars`? How is that different from its type?

**Answer:**

typeof gives the internal storage type, while class is the object oriented behavioral type, so this controls how functions operate on it. 


2. How many observations and variables are in the data?

**Answer:**

```{r}
dim(cars)
```



3. In this data frame, what does one row represent? What does one column represent?

**Answer:**

Each row represents a car, and each column represents some sort of information about that car (variable or characteristic).


## Task 1.2: Use R help

Run `?mtcars` in the Console. Find the definitions and units of the following variables.

- `mpg`:
- `wt`:
- `hp`:
- `am`:

Why would it be risky to analyze a variable without first checking its definition and units?

Without the definition and units, we could misinterpret the magnitude, direction, or real-world meaning of a result. We might also mistake a category code such as `am` for a quantitative variable.

**Answer:**

```{r}
?mtcars
```

```{r}
mean(cars$wt)
```


> **Switch roles.** The navigator becomes the driver.

# Part 2: Vectors, indexing, and functions (15--20 minutes)

The `$` operator selects a column from a data frame. A single column is a vector.

```{r make-vectors}
#this is a scalar
x <- 10

# this is a vector
mpg <- cars$mpg
hp <- cars$hp
```

## Task 2.1: Examine vectors

Write code to answer each question. Use the examples in the introduction notebook if you need a reminder.

1. What is the class of `mpg`?

```{r vector-class}
# YOUR CODE HERE
class(mpg)
typeof(mpg)
```

2. Display the first value in `mpg`.

```{r first-value}
# YOUR CODE HERE
mpg[1]
```

3. Display the fifth through tenth values in `mpg`.

```{r vector-subset}
# YOUR CODE HERE
mpg[5:10]
```

4. Calculate the mean, median, minimum, maximum, and standard deviation of `mpg`.

```{r summary-functions}
# YOUR CODE HERE
mean(mpg)
median(mpg)
min(mpg)
max(mpg)
sd(mpg)
```

5. Create a new vector called `mpg_centered` by subtracting the mean of `mpg` from every value in `mpg`. Then calculate the mean of the new vector.

mpg[1] - mean(mpg)
mpg[2] - mean(mpg)

mpg[n] - mean(mpg)


$$
E[X - \mu] \\
E[X] - \mu \\
\mu - \mu = 0
$$




```{r vector-operation}
# YOUR CODE HERE
mpg_centered <- mpg - mean(mpg)
mean(mpg_centered)
```

Why is the mean not necessarily displayed as exactly zero? If needed, try `round(mean(mpg_centered), 10)`.



**Answer:**

The unrounded result is really close to zero but may not equal zero because computers store a lot of decimal values with finite binary precision. This is ordinary floating-point rounding error.


## Task 2.2: Diagnose an error

The following code contains an error. Run it, read the error message, and fix the code.

```{r error-practice, error=TRUE}
#mean(mpg_data)
```

What did the error message tell you? Write the corrected command below.

```{r error-fix}
# YOUR CORRECTED CODE HERE
mean(mpg)
```

> **Switch roles.** The navigator becomes the driver.

# Part 3: Ask a question with a data frame (20--25 minutes)

We will investigate this question:

> **How does fuel economy differ between cars with automatic and manual transmissions?**

In `mtcars`, `am = 0` indicates an automatic transmission and `am = 1` indicates a manual transmission.

## Task 3.1: Select observations

The code below creates separate data frames for automatic and manual cars.

Read it together before running it. Explain what appears on each side of the comma inside the brackets.

```{r subset-data}
automatic <- cars[cars$am == 0, ]
manual <- cars[cars$am == 1, ]
```

Use `nrow()` to determine how many cars are in each group.

```{r group-counts}
# YOUR CODE HERE
nrow(automatic)
nrow(manual)
```

## Task 3.2: Calculate and interpret group means

Calculate mean `mpg` separately for automatic and manual cars. Give the results clear object names.

```{r group-means}
# YOUR CODE HERE

mean_mpg_automatic <- mean(automatic$mpg)
mean_mpg_manual <- mean(manual$mpg)

mean_mpg_automatic
mean_mpg_manual
```

Report the difference as:

$$
\text{mean MPG for manual cars} - \text{mean MPG for automatic cars}.
$$

```{r mean-difference}
# YOUR CODE HERE
mpg_difference <- mean_mpg_manual - mean_mpg_automatic
mpg_difference
```

Interpret the difference in one sentence. Be precise about its direction and units.

**Answer:**

In this sample, manual cars average about 7.24 more miles per gallon than automatic cars.


Does this comparison establish that changing a car from automatic to manual would *cause* its fuel economy to change by that amount? Why or why not?

**Answer:**

No. Transmission type wasn't randomly assigned, and the two groups may be different in weight, horsepower, model design, engine size, or other characteristics that also affect fuel economy. This is a descriptive association, not necessarily a causal effect.

## Task 3.3: Make and improve a plot

Start with a scatterplot of vehicle weight and fuel economy.

```{r first-plot}
plot(cars$wt, cars$mpg)
```

Modify the plot so that it has:

- a descriptive title;
- informative axis labels that include units;
- filled plotting points (`pch = 19`); and
- a color of your choice.

```{r improved-plot}
# YOUR CODE HERE
plot(
  cars$wt,
  cars$mpg,
  main = "Fuel Economy and Vehicle Weight",
  xlab = "Vehicle weight (thousands of pounds)",
  ylab = "Fuel economy (miles per gallon)",
  pch = 19,
  col = "#A30D26"
)
```

Describe the relationship shown in the plot. Comment on its direction, strength, and whether it appears approximately linear.

**Answer:**

Fuel economy and vehicle weight have a strong negative, approximately linear relationship, heavier cars generally get fewer miles per gallon. A few observations are different from the main pattern, but there isn't any obvious strong curvature.

> **Switch roles.** The navigator becomes the driver.

# Part 4: Write a custom function (15--20 minutes)

A custom function has a name, one or more arguments, actions, and an output. The function below converts miles per gallon to liters used per 100 kilometers.

```{r conversion-function}
mpg_to_l100km <- function(mpg_value) {
  liters_per_100km <- 235.215 / mpg_value
  return(liters_per_100km)
}
```

## Task 4.1: Test and explain the function

1. Predict the result of `mpg_to_l100km(20)`, then run it.

```{r test-function}
# YOUR CODE HERE
mpg_to_l100km(20)
```

2. Use the function on the entire `mpg` vector. Store the result as `liters_per_100km` and display its first six values.

```{r vectorized-function}
# YOUR CODE HERE
liters_per_100km <- mpg_to_l100km(mpg)
head(liters_per_100km)
```

3. In your own words, identify the function's argument, action, and output.

**Answer:**


## Task 4.2: Write your own function

Write a function called `standardize()` that takes a numeric vector `x` and returns

$$
z_i = \frac{x_i - \bar{x}}{s_x}.
$$

```{r standardize-function}
# YOUR CODE HERE
standardize <- function(x) {
  z <- (x - mean(x)) / sd(x)
  return(z)
}
```

Test your function on `hp`. A correctly standardized vector should have a mean of approximately 0 and a standard deviation of 1.

```{r test-standardize}
# YOUR CODE HERE
mean(standardize(hp))
sd(standardize(hp))
```

If your check does not work, use `?mean`, `?sd`, or the error message to help diagnose the problem before asking for help.

# Part 5: Loop and compare (10--15 minutes)

The function in Part 4 works directly on a vector, so a loop is not the most efficient solution here. We will nevertheless use a short loop to practice its logic.

## Task 5.1: Trace a loop

Before running this code, predict what will be stored in `first_five_l100km`.

```{r conversion-loop}
first_five_l100km <- c()

for (i in 1:5) {
  converted_value <- mpg_to_l100km(mpg[i])
  first_five_l100km <- c(first_five_l100km, converted_value)
}

first_five_l100km
```

Explain the role of each component.

- **Loop object (`i`):**
- **Loop vector (`1:5`):**
- **Loop code:**

Compare the loop result with the vectorized command below. Are they the same?

```{r compare-loop}
mpg_to_l100km(mpg[1:5])
```

Why might the vectorized version generally be preferable in R?

**Answer:**

The vectorized version is shorter, easier to read, and usually more efficient because R functions and arithmetic are designed to operate on entire vectors. The loop is still useful for teaching and for things that can't be written naturally as vector operations.



# Part 6: Knit and reflect (5--10 minutes)

Before knitting, check that:

- every code chunk runs from top to bottom;
- every written response is complete;
- your plot has a title and labeled axes; and
- both partners' names appear at the top.

Use **Knit > Knit to HTML**. If knitting gives you an error, read the message, find the named chunk, and run the notebook from the beginning. Remember that knitting starts a fresh R session: it can't use objects that exist only in your console.

## Final reflection

1. What is one R command or idea that became clearer during this activity?

**Answer:**


2. What is one question you still have?

**Answer:**


3. Give one specific example of how your partner helped you solve a problem or understand the code.

**Answer:**


# Optional extensions

Complete these only if your pair finishes early.

## Extension A: A second comparison

Choose one additional variable (`hp`, `wt`, or `qsec`). Compare its mean for automatic and manual cars. Create an appropriate plot and write a two-sentence interpretation.

```{r extension-comparison}
# YOUR CODE HERE
mean_hp_automatic <- mean(automatic$hp)
mean_hp_manual <- mean(manual$hp)
mean_hp_manual - mean_hp_automatic

boxplot(
  hp ~ factor(am, labels = c("Automatic", "Manual")),
  data = cars,
  main = "Horsepower by Transmission Type",
  xlab = "Transmission type",
  ylab = "Gross horsepower",
  col = c("gray80", "#A30D26")
)
```

## Extension B: Improve the function

By default, `mean()` and `sd()` return `NA` if a vector contains a missing value. Modify `standardize()` so it works when `x` contains missing values.

Use `?mean` to identify the needed argument, then test the function on:

```{r extension-missing, eval=FALSE}
x_with_missing <- c(1, 2, NA, 4, 5)
x_with_missing <- c(1, 2, NA, 4, 5)

standardize <- function(x) {
  z <- (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
  return(z)
}

standardize(x_with_missing)
```

## Extension C: Add a fitted line

Use `lm()` to estimate a linear relationship between `mpg` and `wt`, then add the fitted line to your scatterplot with `abline()`. Use `?lm` or `?abline` if needed. Does the line summarize the visual pattern reasonably well?

```{r extension-line}
# YOUR CODE HERE
weight_model <- lm(mpg ~ wt, data = cars)

plot(
  cars$wt,
  cars$mpg,
  main = "Fuel Economy and Vehicle Weight",
  xlab = "Vehicle weight (thousands of pounds)",
  ylab = "Fuel economy (miles per gallon)",
  pch = 19,
  col = "#A30D26"
)
abline(weight_model, col = "#101820", lwd = 2)
```
