---
title: "In-Class Exercises Probability and Stats Review"
output: 
  html_document:
    toc: yes
    toc_depth: 2
    df_print: kable
    highlight: tango
#    includes:
#      in_header: wcag-overrides.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#install.packages('tidyverse')
library(tidyverse)


set.seed(123)
```

# 1. Random Variables & Distributions

Warm‑up (no code)

1. In your own words, what is a random variable?
Variable whose value is determined by some underlying experiment, connects probability with data. Assigns some real number to an outcome in the sample space of a random experiment. 


2. What does it mean to say a variable follows a normal distribution?
Its values are symmetrical, bell shaped, most data are clustered around the mean, extreme values go down equally on both sides


```{r}
# Draw 1,000 observations from a normal distribution
# density function
dnorm(x = c(-1, 0, 1), mean = 0, sd = 1)
# cumulative distribution function (takes a value and gives a percent/ probability)
pnorm(q = 0, mean = 0, sd = 1, lower.tail = TRUE)
# quantile function (takes a percentile and gives a value)
qnorm(p = .5, mean = 0, sd = 1, lower.tail = TRUE)

# random generation
obs <- rnorm(n = 100000, mean = 50, sd = 10)
head(obs)
```


## Task 1: Compute the mean and standard deviation of your observations manually using mean() and sd().

* How close are they to the values we specified (50 and 10)?



```{r}
# calculate the mean and standard deviation
mean(obs)
sd(obs)
```


Discussion:

* Why aren’t the sample values exactly 50 and 10?

We're drawing a random sample, no matter how big it is there will always be an element of randomness/ variability

## Task 2: Visualizing the Distribution

```{r}
# create a histogram of your simulated data

#using base R
#hist(obs)

```

```{r}
# using ggplot (in tidyverse)
obs_data <- data.frame(obs = obs)

obs_plot <- ggplot(obs_data, aes(x = obs)) +
  geom_histogram(
    bins = 30,
    fill = "steelblue",
    color = "white"
  ) +
  labs(
    title = "Distribution of Observations",
    x = "Observed value",
    y = "Frequency",
    alt = paste(
      "Histogram of 100,000 observations drawn from a normal distribution",
      "with a mean of 50 and a standard deviation of 10.",
      "The distribution is bell-shaped and centered near 50."
    )
  ) +
  theme_minimal()

obs_plot
```

Group discussion questions:

* Is the distribution symmetric?
Yes, it does look symmetric

* Where is most of the mass concentrated?
Around 50 :) 


# 2. Probability & Events


Concept check

Suppose:

* $P(A) = 0.4$

* $P(B) = 0.3$

* $P(A \cap B) = 0.1$ 
"intersection / and"
The probability that both A and B happen

* Compute $P(A \cup B)$.
"union/ or"
The probability that either A or B happen


General additive rule: $P(A \cup B) = P(A) + P(B) - P(A \cap B)$

Formula (independent events): $P(A \cap B) = P(A) \times P(B)$

If A and B are mutually exclusive, $P(A \cup B) = P(A) + P(B)$

Formula (dependent events): $P(A \cap B) = P(A) \times P(B|A)$

* Are A and B independent? Why or why not?

```{r}
# P(A or B)
.4 + .3 - .1

# are A and B independent
#P(A and B) 
.4 * .3

```


.12 != .1, so $P(A \cap B)$ are not independent 

If $P(A)$ and $P(B)$ were independent, then $P(A)P(B)$ would equal $P(A \cap B)$ but they don't. 




## Task 3: Empirical Probabilities via Simulation

```{r}
# Roll a die 10,000 times and save the sample
set.seed(123)
rolls <- sample(x = 1:6, size = 10000, replace = TRUE)


# Calculate the probability of rolling a 6 from your sample
# == (test if TRUE)
# != (test if FALSE)
mean(rolls == 6)


```


Next:

* Estimate the probability of rolling an even number from your sample.

* Estimate P(roll ≥ 4).

```{r}
# probability of rolling an even number
# %% 2 
mean(rolls %% 2 == 0)

# probability P(roll >= 4)
mean(rolls >= 4)
```


Discussion:

* Why do these values get closer to the theoretical probabilities as sample size increases?

Law of large numbers: as n increases, observed/ empirical values converge to theoretical probabilities

# 3. Law of Large Numbers

## Task 4: Convergence of Sample Means

Generate two samples, one of size 10 and one of size 5,000.

Each of these samples should draw from the 

```{r}
# Generate increasing sample sizes
n <- seq(10, 5000, by = 10)

means <- sapply(n, function(i) mean(rnorm(i, 50, 10)))
```

```{r}
# using base R to create the plot

#plot(n, means, type = "l",
#main = "Law of Large Numbers",
#xlab = "Sample size",
#ylab = "Sample mean")
#abline(h = 50, col = "red")
```

```{r}
# using ggplot to create the plot
lln_plot <- ggplot(
  data.frame(n = n, means = means),
  aes(x = n, y = means)
) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_hline(
    yintercept = 50,
    color = "red",
    linewidth = 0.8
  ) +
  labs(
    title = "Law of Large Numbers",
    x = "Sample size",
    y = "Sample mean",
    alt = paste(
      "Line plot illustrating the law of large numbers.",
      "As the sample size increases, the sample mean converges toward",
      "the population mean of 50, shown by a red horizontal line."
    )
  ) +
  theme_minimal()

lln_plot

```

Law of large numbers: as n grows, $\bar{x}$ converges to $\mu$ (sample value converges to population value)

Central limit theorem: for a big enough sample, the sampling distribution of the $\bar{x}$ is approximately normal.

Questions

1. What happens to the sample mean as n grows?

Sample mean oscillates pretty wildly at small sample sizes, then stabilizes closer to the population mean at larger sample sizes (LLN)

2. Why does this matter for empirical research?
Large samples give reliable estimates (foundation of stat. inference)


# 4. Central Limit Theorem

## Task 5: Sampling Distribution of the Mean

```{r}
# Draw 1000 sample means from a skewed distribution
sample_means <- replicate(10000, mean(rexp(40)))


# create a histogram of the sample means
# using base R

#hist(sample_means)

```

```{r}
# using ggplot to create the histogram

sample_means_data <- data.frame(sample_means = sample_means)

ggplot(sample_means_data, aes(x = sample_means)) +
  geom_histogram(
    bins = 30,
    fill = "steelblue",
    color = "white"
  ) +
  labs(
    title = "Sampling Distribution of the Sample Mean",
    x = "Sample mean",
    y = "Frequency",
    alt = paste(
      "Histogram of 10,000 sample means, each calculated from 40",
      "observations drawn from an exponential distribution.",
      "The sampling distribution is approximately bell-shaped and",
      "centered near 1."
    )
  ) +
  theme_minimal()
```

Group discussion

* Is the original distribution normal?
No, the original distribution is an exponential distribution


* Is the sampling distribution approximately normal?
The sampling distribution of the mean is approximately normal 


* Why is this result important for inference?
This justifies using normal based inference even when the data aren't normal

CLT: for a bit enough sample size, the sampling distribution of the $\bar{x}$ is approximately normal.




# 5. Descriptive Statistics

## Task 6: Working with Real Data

```{r}
# add the mtcars data to your R session

cars <- mtcars

# looking at the data description
?mtcars
```

Next:

* Compute mean, median, and standard deviation of `mpg`.

* Interpret each statistic.

```{r}
# mean
mean(cars$mpg)

# median
median(cars$mpg)

# sd
sd(cars$mpg)
```



## Task 7: Group Comparisons


```{r}
# Compare the mean mpg for automatic vs manual transmission
# we will use the dplyr package from tidyverse
cars |>
  group_by(am) |>
  summarise(mean_value = mean(mpg, na.rm = TRUE))
```

Questions:

* Which group has higher fuel efficiency?
Manual transmission has higher fuel efficiency



* Is this a causal claim? Why or why not?
Not necessarily (correlation != causation, this is a clue we should dig closer but there could be other elements at play)

# 6. Covariance & Correlation

## Task 8: Correlation in Practice

```{r}
# calculate the correlation between horsepower (hp) and mpg
cor(cars$hp, cars$mpg)
```

Questions:

* Is this relationship positive or negative?

Negative relationship, correlation $\approx -.78$


* How strong is it using the correlation guidelines?

Strong negative relationship



## Task 9: Visualizing Correlation

```{r}
# create a scatterplot with horsepower on the x-axis and mpg on the y-axis
# draw a regression line through the scatterplot

# in base R, to run it remove the "#"

#plot(
#  cars$hp,
#  cars$mpg,
#  main = "Fuel Efficiency and Horsepower",
#  xlab = "Horsepower",
#  ylab = "Miles per gallon",
#  pch = 19,
#  col = "steelblue"
#)

#abline(
#  lm(mpg ~ hp, data = cars),
#  col = "red",
#  lwd = 2
#)

```

```{r}
# in ggplot:

ggplot(cars, aes(x = hp, y = mpg)) +
  geom_point(color = "steelblue") +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    color = "red"
  ) +
  labs(
    title = "Fuel Efficiency and Horsepower",
    x = "Horsepower",
    y = "Miles per gallon",
    alt = paste(
      "Scatterplot of horsepower and fuel efficiency for 32 cars.",
      "The fitted regression line slopes downward, indicating that",
      "cars with greater horsepower tend to have lower fuel efficiency."
    )
  ) +
  theme_minimal()
```

Discussion:

* Does the plot match the correlation sign?

Downward slope matches negative correlation

* What patterns do you see?

Let's check and see if we notice any clustering by transmission type:


```{r}

ggplot(cars, aes(x = hp, y = mpg)) +
  geom_point(
    aes(color = factor(am)),
    size = 2.5
  ) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    color = "red"
  ) +
  scale_color_manual(
    values = c("0" = "steelblue", "1" = "darkorange"),
    labels = c("0" = "Automatic", "1" = "Manual")
  ) +
  labs(
    title = "Fuel Efficiency and Horsepower",
    x = "Horsepower",
    y = "Miles per gallon",
    color = "Transmission",
    alt = paste(
      "Scatterplot of horsepower and fuel efficiency for 32 cars.",
      "Points are colored by automatic or manual transmission.",
      "A single fitted regression line slopes downward, indicating that",
      "cars with greater horsepower tend to have lower fuel efficiency."
    )
  ) +
  theme_minimal()
```

We see that there are some clusters by transmission types.

# 7. Reflection & Wrap‑Up


* **What did simulation help you understand today?**


* **Which concept (LLN, CLT, correlation, etc.) still feels fuzzy?**
Note from me:

These two ideas sound similar, but they answer different questions.

Law of Large Numbers (LLN) asks: “What happens to ONE sample statistic when the sample gets bigger?” As your sample size increases, your sample mean gets closer to the true population mean. Focus: accuracy of a single estimate

Central Limit Theorem (CLT) asks: “What does the distribution of sample means look like across MANY samples?” If you take samples a bunch of times and compute their means, those means form a distribution that becomes approximately normal, even if the original data are skewed. Focus: shape of the sampling distribution

Memory trick:

LLN = one sample, getting closer

CLT = many samples, bell curve

Why both matter: 

- LLN tells us our estimate improves with more data. 

- CLT tells us how that estimate behaves (distribution), which lets us build confidence intervals and hypothesis tests.

* **How could these tools apply to your own research questions?**

Applications to polling, A/B testing, regression, surveys






