---
title: "In-Class Activity Notebook — Nonlinear Regression Functions"
output:
  html_document:
    toc: yes
    toc_depth: 2
    df_print: kable
    highlight: tango
#    includes:
#      in_header: wcag-overrides.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
#library(here)
#here::i_am("in-class-notebooks/inclass_nonlinear.Rmd")
```

```{r}
need <- c("tidyverse", "wooldridge", "sandwich", "lmtest", "car", "modelsummary", "kableExtra")
have <- need %in% rownames(installed.packages())
if (any(!have)) install.packages(need[!have])
invisible(lapply(need, library, character.only = TRUE))
#source(here("scripts", "reporting.R"))
source("reporting.R")
```


# Goals for today

1. Diagnose when a linear regression function is not the best choice and propose a better functional form.
2. Fit and interpret polynomial models (especially quadratics) and compute a turning point.
3. Fit and interpret log models (linear–log, log–linear, log–log).
4. Fit and interpret interaction models, (including interactions with nonlinear terms if we have time)
5. Compare models with adjusted $R^2$ and interpretability.


---

# 1. Load and understand the data

We will use `benefits` from the `wooldridge` package (school-level data).

To use the built-in R summaries as opposed to the summary from `reporting.R` (also `glimpse`, which is from `tidyvserse`), remove the hash marks from in front of those lines of code. 

```{r, results = "asis"}
data("benefits", package = "wooldridge")
df <- benefits
#summary(df)
#glimpse(df)

mk_dataset_summary(df)
```

## 1.1 Variables

Let's take a look at the dataset:

```{r}
?benefits
```


For today, we’ll treat:

- Outcome variable: `avgben` (average teacher non-salary benefits, $)
- Candidate predictors: `avgsal`, `lunch`, `enroll`, `staff`, `exppp`
- Some useful transformations already there: `lavgsal`, `lenroll`, `lstaff`, and `bs = avgben/avgsal`

**Question:** In 2–3 sentences, describe what `lunch` measures and what kind of confounding you worry about in a benefits–salary relationship.

**Answer:**

---

# 2. Is linearity plausible?

## 2.1 Scatterplot: benefits vs salary

Use ggplot to create a scatterplot of average salary (`avgsal`) vs. average non-salary benefits (`avgben`). Put `avgsal` on the x-axis and `avgben` on the y-axis.

```{r, fig.alt = "Scatter plot of average salary vs. average non-salary benefits."}
# create your scatterplot here

```

**Question:** Does a straight line look reasonable? If not, what kinds of nonlinearity do you see (curvature, diminishing returns, etc.)?


**Answer:**


---

## 2.2 Baseline linear model

Create a linear model:

$avgben = \beta_0 + \beta_1 avgsal + \beta_2 lunch + \beta_3 enroll + \beta_4 staff + \beta_5 exppp + u$. 

Look at the model results.

```{r, results = "asis"}
# create your linear model

# look at the model results

```

**Question:** Interpret the coefficient on `avgsal` as a *ceteris paribus* effect. How does this interpretation square with what you saw in your scatterplot? 

**Answer:**

---

# 3. Log models: three common cases

We will work with these families:

- **Linear–log:** $Y = \beta_0 + \beta_1 \ln(X) + u$
- **Log–linear:** $\ln(Y) = \beta_0 + \beta_1 X + u$
- **Log–log:** $\ln(Y) = \beta_0 + \beta_1 \ln(X) + u$

## 3.1 Check positivity before logging

**Question**: Why is it important to check positivity before logging?


**Answer**:


---



## 3.2 Linear–log: benefits on log(salary)

Create a linear-log model:

$avgben = \beta_0 + \beta_1 ln(avgsal) + \beta_2 lunch + \beta_3 ln(enroll) + \beta_4 ln(staff) + \beta_5 exppp + u$. 

Look at the model results.

```{r}
# create your linear-log model

# look at your model results


```


**Question:** Interpret $\hat\beta_1$ on `ln(avgsal)` in a linear–log model. How is this different from the level model? Is this easier to interpret? Harder? Why?


**Answer:**


---

## 3.3 Log–log: log(benefits) on log(salary)

Create a log-log model:

$ln(avgben) = \beta_0 + \beta_1 ln(avgsal) + \beta_2 lunch + \beta_3 ln(enroll) + \beta_4 ln(staff) + \beta_5 exppp + u$. 

Look at the model results.

```{r}
# create your log-log model

# look at your model results


```

**Question:** Interpret $\hat\beta_1$ on `lavgsal` as an elasticity. What is an elasticity? Which (out of the three models) do you prefer from an interpretation standpoint?

**Answer:**



---

# 4. Polynomial models 

A quadratic regression function is:


$E[Y|X] = \beta_0 + \beta_1 X + \beta_2 X^2$


## 4.1 Quadratic in salary

Create a model that is quadratic in salary: 


$avgben = \beta_0 + \beta_1 avgsal + \beta_2 avgsal^2 + \beta_3 lunch + \beta_4 enroll + \beta_5 staff + \beta_6 exppp + u$. 

Look at the model results.


```{r, results = "asis"}
# create your quadratic model

# look at your model results


```

**Question:** Why is it incorrect to include $X^2$ without $X$?

**Answer:**


---



## 4.2 Turning point + inside/outside the data range

Compute the turning point $-\beta_1/(2\beta_2)$. Then compare it to the observed salary range.

```{r}
# compute the turning point

```


**Question:** Is the turning point inside the observed range of `avgsal`? Does that matter for interpretation? Why or why not? 


**Answer:**


---



## 4.3 RESET test

RESET tests whether functions of the fitted values add explanatory power. 

So after fitting a polynomial model, RESET checks whether higher-order nonlinear combinations of the prediction still matter.

If yes, your polynomial may be misspecified (like the degree too low).

Use `?resettest` to look at the documentation for the RESET Test.

```{r}
?resettest
```


First, fit a simplified quadratic model:

$avgben = avgsal + avgsal^2 + u$

And fit a RESET test up to the cubed value of $avgsal$:

```{r}
# fit your RESET test here

```

**Question:** What does the `power` argument mean in the RESET test? What is it testing? What test statistic is it using?


**Answer:**




---



# 5. Interactions 

An interaction lets the effect of one variable depend on another.

We’ll start with an interaction between salary and poverty exposure (`lunch`).

$avgben = \beta_0 + \beta_1 avgsal \cdot lunch + \beta_2 enroll + \beta_3 staff + \beta_4 exppp + u$. 

Create your model and look at your results.

## 5.1 Interaction: salary × lunch

```{r, results = "asis"}
# create your interaction model

# look at your model results

```

**Question:** Write down the fitted regression function $\widehat{E[avgben|avgsal,lunch,\dots]}$ and the partial effect
$\partial \widehat{E[avgben|\cdot]}/\partial avgsal$. How does it depend on `lunch`?


**Answer:**


---

## 5.2 Interaction + curvature (quadratic × lunch)

Now allow salary to be nonlinear **and** allow the relationship to differ with lunch.


$avgben = \beta_0 + \beta_1 [(avgsal_c \cdot avgsal_c^2) \cdot lunch] + \beta_2 enroll + \beta_3 staff + \beta_4 exppp + u$. 

Create your model and look at your results.

```{r, results = "asis"}
# create your nonlinear interaction model here

# look at your model summary


```

**Question:** Is there evidence that the salary–benefits relationship differs by `lunch` *and* varies nonlinearly? Point to the specific coefficients you used as evidence.

**Answer:**



---

# 6. Compare models: fit *and* interpretability

Compare:
1. Linear (no interaction)
2. Linear with interaction
3. Quadratic (no interaction)
4. Quadratic with interaction

```{r}
# display your linear model results

# display your linear model with interaction results

# display your quadratic model without interaction results

# display your quadratic model with interaction results



```



**Question:** Which model would you recommend to a policymaker, and why? Your answer should emphasize interpretation and plausibility, not just fit.


**Answer:**





---

# 7. Prediction + interpretation check

Create a hypothetical school and produce predicted benefits under two models.

Assign plausible values to your new school.

```{r}
#new_school <- tibble(
#  avgsal = # assign a value for average salary here,
#  lunch  = # assign a value for lunch here,
#  enroll = # assign a value for enrollment here,
#  staff  = # assign a value for staff here,
#  exppp  = # assign a value for expenditures here
#) %>%
#  mutate(avgsal_c = avgsal - mean(df$avgsal, na.rm = TRUE))

#new_school
```


Make your predictions using your two favorite models from task 6 and your new_school predictions dataframe:

```{r}
# make your benefits predictions here
```

**Question:** In 2–3 sentences: are these predictions sensible? What’s one limitation of predicting from a model that includes polynomials/interactions?


**Answer:**

---

# Wrap-up 

Before you leave, write a 3–5 sentence reflection:

1. What regression function did you prefer?
2. What did you learn about why we use logs/polynomials/interactions?
3. What would you check next if this were a real policy brief?


