---
title: "ECON-8300"
output: 
  html_document:
    df_print: kable
    highlight: tango
#    includes:
#      in_header: wcag-overrides.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Homework03

Grading Rubric (per question):  
2 points if complete and correct  
0 points otherwise 


## Retrieving your work

This and all future homework .Rmd files are written in Markdown. The .Rmd file can be rendered as an .html file for easier viewing. 

Once you have opened the .Rmd file, you may begin writing your answers beneath each problem. Do not remove any of the original problems from the file, because that makes it more difficult for the me to grade your assignment. 

## Submitting your work

When you upload your files to Canvas, you are to submit ***two*** files:

a. Your reproducible document file (.Rmd) which should be saved as homework##-ID.Rmd. For example, my homework03 file would be saved as homework03-jperrigan.Rmd.

b. Your rendered reproducible document file (.html) which should be saved as homework##-ID.html. For example, my homework03 file would be saved as homework03-jperrigan.html.

Your ID is your Canvas username. It is also the text before the "@" symbol in your UNO email address.

You have an unlimited number of submissions, but only the latest submission will be viewed and graded. **Proper means that the file is correctly named and formatted. Latest means that the file is the most recent submission prior to the deadline. Remember the .Rmd file needs to render properly to .html before submitting.** 


***

All students are expected to complete all problems. If any problem asks you a particular question, be sure to answer it completely (with code, written sentences, or both).

***Do not change anything in this file above the double line. Doing so will prevent your homework from being graded.***

***
***

# Data and packages

For the first part of the assignemnt, we will use the `meap01` dataset from the `wooldridge` package. It contains data for 1,823 Michigan elementary schools. The variables used here are:

- `math4`: percentage of fourth-grade students rated satisfactory in mathematics
- `exppp`: expenditures per pupil, in dollars
- `lexppp`: natural logarithm of expenditures per pupil
- `lunch`: percentage of students eligible for free or reduced-price lunch
- `enroll`: school enrollment

```{r packages}
# Install the package once if needed:
# install.packages("wooldridge")

library(wooldridge)
data("meap01")
```


***

**#1** 

**Visualizing and quantifying the relationship**

1. Construct a scatterplot with `exppp` on the horizontal axis and `math4` on the vertical axis. Add a fitted linear regression line and informative axis labels.
2. In 2–3 sentences, describe the direction, strength, and shape of the relationship. Note any observations that might have unusually high influence. Is there a high correlation by sight? What about when you calculate the correlation?

Save your written response as `prob01`.

**Answer:**

```{r, problem01}
# put your plot code here
# prob01 <-
```


***

**#2** 

Consider the population regression model

$$
math4_i = \beta_0 + \beta_1 exppp_i + u_i.
$$

In 4–6 sentences:

1. Explain the economic meaning of $\beta_0$ and $\beta_1$, including their units. Are they population parameters or sample estimates?
2. Explain what the error term $u_i$ contains. Give two concrete examples of factors that could belong in $u_i$.

Save your response as a string named `prob02`.

**Answer:**

```{r, problem02}
# prob02 <-
```

***


**#3** 

**Calculating the OLS estimators by hand**

Let $x_i=exppp_i$ and $y_i=math4_i$.

1. Compute the sample means $\bar{x}$ and $\bar{y}$.
2. Create the centered variables $x_i^c=x_i-\bar{x}$ and $y_i^c=y_i-\bar{y}$; save them as `xc` and `yc`.
3. Verify numerically that $\sum_i x_i^c=0$ and $\sum_i y_i^c=0$.
4. Calculate

$$
\hat\beta_1=\frac{\sum_i(x_i-\bar{x})(y_i-\bar{y})}{\sum_i(x_i-\bar{x})^2}
$$

and

$$
\hat\beta_0=\bar{y}-\hat\beta_1\bar{x}.
$$

Save the estimates as `b1_hat` and `b0_hat`. Do not use `lm()` in this problem.

**Answer:**

```{r, problem03}
# xc <-
# yc <-
# b1_hat <-
# b0_hat <-
```

***

**#4** 

Estimate

$$
math4_i = \beta_0 + \beta_1 exppp_i + u_i
$$

using `lm()` and save the fitted model as `prob04_model`.

1. Check that the estimated intercept and slope match your calculations in Problem 3 (small differences from rounding are acceptable).
2. Interpret the slope in context. What would be the predicted change in `math4` associated with a $500 increase in expenditures per pupil?

Save the written responses to parts 2 and 3 as a string to `prob04`.

**Answer:**

```{r, problem04}
# prob04_model <-
# prob04 <-
```

***

**#5** 

For `prob04_model`, calculate the following manually from `math4`, `yhat`, and `uhat`:

$$
TSS=\sum_i(y_i-\bar{y})^2, \qquad
ESS=\sum_i(\hat{y}_i-\bar{y})^2, \qquad
SSR=\sum_i\hat{u}_i^2,
$$

$$
R^2=1-\frac{SSR}{TSS}, \qquad
SER=\sqrt{\frac{SSR}{n-2}}.
$$

1. Save these five values in a named vector called `prob05_fit`. Put them in the order: $TSS, ESS, SSR, R^2, SER$.
2. Interpret $R^2$ and the SER in context. Be precise about their units. Make sure your $R^2$ matches the value reported by your model `summary()`.

Save your interpretation as `prob05`.

**Answer:**

```{r, problem05}
# prob05_fit <-
# prob05 <-
```

***


**#6** . 

The OLS slope can be written as

$$
\hat\beta_1=\beta_1+
\frac{\sum_i(x_i-\bar{x})u_i}{\sum_i(x_i-\bar{x})^2}.
$$

In 6–8 sentences:

1. Use this expression to explain why $E(u_i\mid X_i)=0$ implies that $\hat\beta_1$ is unbiased.
2. Explain why $\hat\beta_1$ varies from sample to sample even when $\beta_1$ is fixed.
3. Explain how increasing the sample size generally affects the sampling distribution of $\hat\beta_1$.
4. Distinguish unbiasedness from consistency.

Save your response as a string named `prob06`.

**Answer:**

```{r, problem06}
# prob06 <-
```


***

# Data and packages

For the rest of the homework, we will use the `bwght` dataset from the `wooldridge` package. It contains data on births in Pennsylvania. The variables used here are:

- `bwght`: infant birth weight, in ounces
- `smoker`: indicator equal to 1 if the mother smoked during pregnancy and 0 otherwise
- `cigs`: average number of cigarettes smoked per day during pregnancy
- `faminc`: family income, in thousands of dollars
- `parity`: birth order of the child

Use homoskedasticity-only standard errors unless otherwise specified.

```{r packages}
# Install these packages once if needed:
# install.packages(c("sandwich", "lmtest"))

library(sandwich)
library(lmtest)
data("bwght")
bwght$smoker <- ifelse(bwght$cigs > 0, 1, 0)
```

**#7** 

1. Compute the mean birth weight for nonsmokers (`smoker = 0`) and smokers (`smoker = 1`). Save the two means in a named vector called `prob07_means`.
2. Estimate

$$
bwght_i=\beta_0+\beta_1 smoker_i+u_i
$$

and save the model as `prob07_model`.
3. Show numerically how the two regression coefficients are related to the two group means.
4. Interpret the intercept and slope in context and include their units. 

Save your written response as `prob07`.

**Answer:**

```{r, problem07}
# prob07_means <-
# prob07_model <-
# prob07 <-
```

***

**#8** 

For the two smoking groups:

1. Compute each group's sample size and sample variance of `bwght`.
2. Manually calculate the standard error of the difference in sample means and save this standard error as `prob08_se`:

$$
SE(\bar{Y}_1-\bar{Y}_0)
=\sqrt{\frac{s_1^2}{n_1}+\frac{s_0^2}{n_0}}.
$$

$$
H_0:\beta_1=0
\qquad \text{against} \qquad
H_1:\beta_1\neq 0.
$$

3. Using the calculated standard error and the $\hat{\beta}_1$ from the model in problem 7, calculate the test statistic and its associated two-sided large-sample p-value manually for the test above and save the p-value as `prob08_p`. (Hint: use ```pnorm()```).
4. State the decision at the 5% significance level and interpret the evidence in context.

Save your written conclusion as `prob08`.

**Answer:**

```{r, problem08}
# prob08_se <-
# prob08_t <- 
# prob03_p <-
# prob03 <-
```


***

**#9** 

Test whether smoking is associated with **lower** average birth weight:

$$
H_0:\beta_1\geq 0
\qquad \text{against} \qquad
H_1:\beta_1<0.
$$

1. Calculate the one-sided large-sample p-value using the t-statistic from Problem 8; save it as `prob09_p`.
2. State the decision at the 1% significance level.
3. Explain why this p-value is different from the two-sided p-value and why the direction of the alternative hypothesis must be picked before looking at the estimate.

Save your explanation as `prob09`.

**Answer:**

```{r, problem09}
# prob09_p <-
# prob09 <-
```


***


**#10**  

Answer the following in 7–9 sentences:

1. What is the sampling distribution of an OLS estimator?
2. What does its standard error measure?
3. What does the central limit theorem contribute to large-sample regression inference?
4. Why does Stock and Watson emphasize heteroskedasticity-robust standard errors in observational economic data?
6. Distinguish statistical significance from economic importance, using the smoking coefficient as an example.

Save your response as a string named `prob10`.

**Answer:**

```{r, problem10}
# prob10 <-
```

***
